State of the Practice for Lattice Boltzmann Method Software

### 2022-03-07 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
digits under

    We analyze the state of software development practice for Lattice Boltzmann
    solvers by quantitatively and qualitatively measuring and comparing 24
    software packages for 10 software qualities (installability, correctness/
    verifiability, reliability, robustness, usability, maintainability,
    reusability, understandability, visibility/transparency and
    reproducibility). Our reproducible analysis method employs a measurement
    template (containing 108 measures that are manually and automatically
    extracted from the project repository) and developer interviews (consisting
    of a set of 20 questions). From the measurement results, we ranked the
    software using the Analytic Hierarchy Process (AHP). Our ranking was roughly
    consistent with ranking the repositories by the number of GitHub stars,
    although the number one project by star count (Sailfish) is ranked 9th by
    our methodology. Our finding is that the state of the practice is healthy
    with 67\% of the measured packages ranking in the top 5 for at least one
    quality, the majority of LBM generated artifacts corresponding to general
    recommendations from research software developers, common utilization of
    version control (67\% of packages) and adoption of a quasi-agile development
    process.  Areas of best practice to potentially improve include adoption of
    continuous integration, API documentation and enforcement of programming
    style guides.  We interviewed four developers to gain insight into their
    current pain points. Identified challenges include lack of development time,
    lack of funding, and difficulty with ensuring correctness.  Developers are
    addressing these pain points by designing for change, circumventing the
    oracle problem and prioritizing documentation and usability.  For future
    best practices, we suggest the community consider the following: employing
    linters, conducting rigorous peer reviews, writing and submitting more papers on software, growing the number of contributors and augmenting the theory manuals to include more requirements relevant information.

@@

We rank the projects by how well they follow best practices
(\rqref{RQ_HighestQuality}) via a measurement template, as described in
\citet{SmithEtAl2021}.  For each software package (each column in the template),
we fill-in the rows of the template. This process takes about 2 hours per
package, with a cap of 4 hours. The time constraint is necessary so that the
work load is feasible for a team as small as one, given our aim to cap the
measurement phase at 160 person hours \citep{SmithEtAl2021}.  An excerpt of the
template, in spreadsheet form, is shown in
Figure~\ref{measurement_template_image}.

@@

The full template consists of 108 questions categorized under 9 qualities:
\begin{inparaenum}[(i)]
	\item installability;
	\item correctness and verifiability;
	\item surface reliability;
	\item surface robustness;
	\item surface usability;
	\item maintainability;
	\item reusability;
	\item surface understandability; and,
	\item visibility/transparency. 
\end{inparaenum}

@@

We requested interviews with developers for all 24 packages, except Munubi,
since Munubi was added after our ethics approved had expired.  We were able to
recruit 4 developers to participate in our study.  Our potential interview
subjects were found from project websites, code repositories, publications, and
biographic information on institution web-pages. Meetings were held using on-line meeting software (either Zoom or Teams), which facilitated recording and
automatic transcription.  Each meeting lasted about an hour. The interview
process was approved by the McMaster University Research Ethics Board under the
application number 
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.

@@

Although the software packages all apply LBM to solve scientific computing
problems, the packages focus on varied CFD problems, with varying parameters,
and are technically different from each other (as shown in
Table~\ref{tbl_features}). Due to this, a comparison of run-time performance is
not appropriate. Instead we looked through each software package's artifacts for
evidence that performance was considered. The artifacts of 18 software packages
mentioned parallelization. This included GPU processing and the CUDA parallel
computing platform, which were mentioned in the artifacts of 6 packages
(ESPResSo, lbmpy, lettuce, pyLBM, Sailfish, TCLB). When a high degree of
parallelization is possible GPUs provide superior speed compared to CPUs. The
software package TCLB is implemented in a highly efficient multi-GPU code to
achieve performance suitable for model optimization \citep{rutkowski2020open}.
In the Ludwig package, a so-called mixed mode approach is used where
fine-grained parallelism is implemented on the GPU, and MPI is used for even
larger scale parallelism \citep{gray2013ludwig}. While one software package
(Sailfish) required CUDA and GPU processing, some (ESPResSo, lbmpy, lettuce,
pyLBM, TCLB) have the option of using either the GPU or the CPU. In general, the
packages that require GPU and CUDA have better performance at the expense of
installability and surface reliability.

@@

Figure~\ref{Fig_Maintainability} shows the maintainability ranking of the
software packages. Software packages with a high score provide version numbers
on documents and source code releases, have an abundance of high quality
artifacts, and use an issue tracking tool and version control system. These
packages also appear to manage their issue tracking, having most of their issues
closed. Their code files are well commented with more than 10 percent of the
code being comments. Four of the top 5 ranked packages are noted as being alive.

@@

Our methodology involved the following steps: i) Identify a list of 45 candidate
software packages; ii) Filter the list to a length of 24 packages; iii) Collect
repository related data on each software package, like number of stars, number
of open issues, number of lines of code; iv) Fill in the measurement template
(the template consists of 108 questions to assess 9 qualities (installability,
correctness/verifiability, reliability, robustness, usability, maintainability,
reusability, understandability, and visibility/transparency).  Filling in the
template requires installing the software on a VM, running simple tests (like
completing the getting started instructions (if present)), and searching the
code, documentation and test files; v) Rank the software using the Analytic
Hierarchy Process (AHP); vi) Interview developers (the interview consists of 20
questions and takes about an hour); and, vii) Conduct a domain analysis. The
collected data was analyzed by: a) comparing the ranking by best practices
against the ranking by popularity; b) comparing artifacts, tools and processes
to current research software development guidelines; and, c) exploring pain
points.  Our ranking was roughly consistent with ranking the repositories by the
number of GitHub stars, although the number one project by star count (Sailfish)
is ranked 9th by our methodology.

Current LBM software provides many examples of following best practices, to the
benefit of users and developers.  This is illustrated by
Table~\ref{topperformerstable}, which summarizes the top 5 performers for each of
the identified qualities and for the overall quality. Taking the union of all
entries in the table, we have 16 top performers: ESPREsSo, MechSys, OpenLB,
DL\_MESO, Sailfish, Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce,
ESPREsSo++, LUMA, MP-LABS, Palabos, and HemeLB.  Therefore, two thirds (67\%) of
the 24 measured packages appear in the top 5 for at least one quality.  Of these
entries almost all of them are classified as alive, except for three packages:
laboetie, MP-LABS, and HemeLB.  This suggests that alive packages are in a
healthy state with respect to adopting best practices.

>>>>

    We analyze the state of software development practice for Lattice Boltzmann
    solvers by quantitatively and qualitatively measuring and comparing 24
    software packages for 10 software qualities (installability, correctness/
    verifiability, reliability, robustness, usability, maintainability,
    reusability, understandability, visibility/transparency and
    reproducibility). Our reproducible analysis method employs a measurement
    template (containing 108 measures that are manually and automatically
    extracted from the project repository) and developer interviews (consisting
    of a set of 20 questions). From the measurement results, we ranked the
    software using the Analytic Hierarchy Process (AHP). Our ranking was roughly
    consistent with ranking the repositories by the number of GitHub stars,
    although the number one project by star count (Sailfish) is ranked 9th by
    our methodology. Our finding is that the state of the practice is healthy
    with 67\% of the measured packages ranking in the top five for at least one
    quality, the majority of LBM generated artifacts corresponding to general
    recommendations from research software developers, common utilization of
    version control (67\% of packages) and adoption of a quasi-agile development
    process.  Areas of best practice to potentially improve include adoption of
    continuous integration, API documentation and enforcement of programming
    style guides.  We interviewed four developers to gain insight into their
    current pain points. Identified challenges include lack of development time,
    lack of funding, and difficulty with ensuring correctness.  Developers are
    addressing these pain points by designing for change, circumventing the
    oracle problem and prioritizing documentation and usability.  For future
    best practices, we suggest the community consider the following: employing
    linters, conducting rigorous peer reviews, writing and submitting more papers on software, growing the number of contributors and augmenting the theory manuals to include more requirements relevant information.

@@

We rank the projects by how well they follow best practices
(\rqref{RQ_HighestQuality}) via a measurement template, as described in
\citet{SmithEtAl2021}.  For each software package (each column in the template),
we fill-in the rows of the template. This process takes about two hours per
package, with a cap of four hours. The time constraint is necessary so that the
work load is feasible for a team as small as one, given our aim to cap the
measurement phase at 160 person hours \citep{SmithEtAl2021}.  An excerpt of the
template, in spreadsheet form, is shown in
Figure~\ref{measurement_template_image}.

@@

The full template consists of 108 questions categorized under nine qualities:
\begin{inparaenum}[(i)]
	\item installability;
	\item correctness and verifiability;
	\item surface reliability;
	\item surface robustness;
	\item surface usability;
	\item maintainability;
	\item reusability;
	\item surface understandability; and,
	\item visibility/transparency. 
\end{inparaenum}

@@

We requested interviews with developers for all 24 packages, except Munubi,
since Munubi was added after our ethics approved had expired.  We were able to
recruit four developers to participate in our study.  Our potential interview
subjects were found from project websites, code repositories, publications, and
biographic information on institution web-pages. Meetings were held using on-line meeting software (either Zoom or Teams), which facilitated recording and
automatic transcription.  Each meeting lasted about an hour. The interview
process was approved by the McMaster University Research Ethics Board under the
application number 
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.

@@

Although the software packages all apply LBM to solve scientific computing
problems, the packages focus on varied CFD problems, with varying parameters,
and are technically different from each other (as shown in
Table~\ref{tbl_features}). Due to this, a comparison of run-time performance is
not appropriate. Instead we looked through each software package's artifacts for
evidence that performance was considered. The artifacts of 18 software packages
mentioned parallelization. This included GPU processing and the CUDA parallel
computing platform, which were mentioned in the artifacts of six packages
(ESPResSo, lbmpy, lettuce, pyLBM, Sailfish, TCLB). When a high degree of
parallelization is possible GPUs provide superior speed compared to CPUs. The
software package TCLB is implemented in a highly efficient multi-GPU code to
achieve performance suitable for model optimization \citep{rutkowski2020open}.
In the Ludwig package, a so-called mixed mode approach is used where
fine-grained parallelism is implemented on the GPU, and MPI is used for even
larger scale parallelism \citep{gray2013ludwig}. While one software package
(Sailfish) required CUDA and GPU processing, some (ESPResSo, lbmpy, lettuce,
pyLBM, TCLB) have the option of using either the GPU or the CPU. In general, the
packages that require GPU and CUDA have better performance at the expense of
installability and surface reliability.

@@

Figure~\ref{Fig_Maintainability} shows the maintainability ranking of the
software packages. Software packages with a high score provide version numbers
on documents and source code releases, have an abundance of high quality
artifacts, and use an issue tracking tool and version control system. These
packages also appear to manage their issue tracking, having most of their issues
closed. Their code files are well commented with more than 10 percent of the
code being comments. Four of the top five ranked packages are noted as being alive.

@@

Our methodology involved the following steps: i) Identify a list of 45 candidate
software packages; ii) Filter the list to a length of 24 packages; iii) Collect
repository related data on each software package, like number of stars, number
of open issues, number of lines of code; iv) Fill in the measurement template
(the template consists of 108 questions to assess nine qualities (installability, correctness/verifiability, reliability, robustness, usability, maintainability, reusability, understandability, and visibility/transparency).  Filling in the template requires installing the software on a VM, running simple tests (like completing the getting started instructions (if present)), and searching the code, documentation and test files; v) Rank the software using the Analytic Hierarchy Process (AHP); vi) Interview developers (the interview consists of 20 questions and takes about an hour); and, vii) Conduct a domain analysis. The collected data was analyzed by: a) comparing the ranking by best practices against the ranking by popularity; b) comparing artifacts, tools and processes to current research software development guidelines; and, c) exploring pain points.  Our ranking was roughly consistent with ranking the repositories by the number of GitHub stars, although the number one project by star count (Sailfish) is ranked 9th by our methodology.

Current LBM software provides many examples of following best practices, to the
benefit of users and developers.  This is illustrated by
Table~\ref{topperformerstable}, which summarizes the top five performers for each of the identified qualities and for the overall quality. Taking the union of all entries in the table, we have 16 top performers: ESPREsSo, MechSys, OpenLB, DL\_MESO, Sailfish, Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce,
ESPREsSo++, LUMA, MP-LABS, Palabos, and HemeLB.  Therefore, two thirds (67\%) of
the 24 measured packages appear in the top five for at least one quality.  Of these entries almost all of them are classified as alive, except for three packages: laboetie, MP-LABS, and HemeLB.  This suggests that alive packages are in a healthy state with respect to adopting best practices.
